Refunctionalization of Abstracting Abstract Machines (Functional Pearl)

### 2018-03-16 paper.tex
misc

%% Title information
%\title{Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)}         %% [Short Title] is optional;
\title{Refunctionalization of Abstracting Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap Between Abstracting Abstract Machines and Abstracting Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and monadification (or un-CPS to direct-style). Linearization expresses
  non-deterministic choices into first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

\subsection{Contributions}

@@

We summarize our contributions as follows:

\begin{itemize}
\item The main contribution of this paper is the filling of the gap
  between small-step AAM and big-step ADI by developing a series of
  systematic transformations.
  We show that the correspondence between abstract
  machines and interpreters exists not only in concrete semantics artifact,
  but also in abstract semantics artifact.
  Galois connection?
  \note{JD: Section number?}

\item A new understanding of pushdown control-flow analysis by
  refunctionalization.
  We present an abstract interpreter obtained from refunctionalizing
  a small-step abstracting abstract machine.
  This abstract interpreter naturally exhibits the pushdown control-flow property,
  as the refunctionalization forces us to sequentialize the
  abstract execution.
  We also analyze its correctness, soundness, and termination.

\item Further, we show that defunctionalization and refunctionalization
  of abstract interpreters play important roles for the call stack of the analyzed language.
  Given this insight, we reveal precisely \emph{why} small-step AAM has no pushdown
  control-flow property (without extra efforts), as well as explain
  why ADI naturally extends pushdown control-flow property from its
  defining language.

\subsection{Outline}

\note{TR: can this be merged with contributions? (they currently lack section refs)}
The remainder of this paper is organized as follows.
We begin by describing the target language we will analyze in Section~\ref{background},
including a concrete CESK machine as small-step operational semantics for it.
We continue with a review of abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines. Following this, Section~\ref{linear} shows a conversion from the
nondeterministic choices of closures to another layer of continuations,
therefore linearizing the abstract execution.
Section~\ref{fusing} then discusses fusing the driver and step functions to a single
function.
We further identify in Section~\ref{disen} the first-order data type representations of
continuations, and disentangle the code blocks for dispatching data types to separate
functions.
Section~\ref{refunc} then represents the refunctionalized abstract abstract machines.
In this section, the continuations represented by data types become higher-order
functions.
In Section~\ref{directstyle}, we transform the refunctionalized AAM to two
different direct style, one is implemented with monads, the other is using
delimited controls.

Finally, we discuss related work in Section~\ref{sec:related}, followed by concluding thoughts in Section~\ref{sec:conclusion}.

\todo{Add section numbers to the figure}\note{TR: put this into a proper figure, add known danvy sequence 
for concrete evaluators at the bottom (in grey or dashed lines)}

@@

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use $\lambda$-calculus in administrative normal form (ANF)
which is a direct-style intermediate representation as target for clarity
but without losing simplicity and generality.
Although we just show the core calculus language, it can be easily extended
to support recursive bindings (such as @letrec@), conditionals, primitive types and
operations on primitive types. These cases would be trivial to implement,
so we omit them in the paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain direct-style $\lambda$-calculus languages.

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language
in ANF \cite{flanagan1993essence}.

@@

In ANF, an expression is an atomic expression or a @let@ expression.
The restriction is that all the function applications must be administrated in a @let@ expression,
and then bound to a variable name under the current environment.
Both the operator and operand of function applications are an atomic expressions.
The atomic expressions $ae$ are either a variable, or a literal @lambda@ term, which
both can be evaluated in a single step.
We also assume that all the variable names in the program are unique.

@@

The CESK machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
The CESK machine models program execution as state transitions in a small-step fashion.
As its name suggested, a machine state has four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program. It maps from addresses to values.
The addresses space is numbers starting from 0.
In our toy language, the only category of value is the closure, i.e., a function paired with an environment.
4) \textit{Continuation} represents the program stack. \todo{explain more on stack}

We show the Scala representations for the components of CESK machine.

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames.
The frame can be considered as evaluation context in reduction semantics.
The frame class @Frame@ stores the information of a single call-site,
i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes of a variable name @x@ to be bind later, a control expression
that resumes to, and its environment. (TODO: mention that frame list vs KArg/KFun)

\subsubsection{Single-step Transition}
Before going into describing how the machine evaluates expressions, we firstly define several helper functions.
As we mentioned in section \ref{anfsyntax}, the atomic expressions are either a variable, or
a literal @lambda@ term. So the atomic expression evaluator @atomicEval@ handles these two
cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates unique integer
in the domain of store.
The @isAtomic@ function is used as a predicate of the expression is atomic.

@@

Now, we can faithfully describe the state transition function @step@,
that given a machine state, determines its successor state.

@@

In the first case, the control of current state is a @Let@ expression,
and its right-hand side is an function application.
By calling evaluator @atomicEval@, we obtain the closure that callee @f@ stands for.
The successor state's control then transfers to the @body@ expression of closure,
with an updated environment and an update store. The new environment is extended
from closure's environment and mapped @v@ to a fresh address @addr@.
The new store is extended with @addr@ mapping to the value of @ae@,
which is evaluated from @atomicEval@.
In the meanwhile, a new frame @frame@ is pushed onto the stack @k@.
This @frame@ contains the variable name @x@ at left-hand side position of @Let@,
the body expression of @Let@, and the lexical environment of the body expression.

The second case for @step@ is the control is transferred to an atomic expression.
We extract the top frame of continuations firstly.
The control (that is an atomic expression) of current state is the evaluated term
that being bind to the variable @x@ from the top frame.
The environment and store are updated with @x@ mapping to the closure value of @ae@.
Then the successor state is transferred to expression @e@ from the top frame,
which is the body of a @Let@ expression, with the updated environment, store, and
the rest of stack @ks@.

@@

To run the program, we first use function @inject@ to construct an initial machine
state given an expression @e@. The initial state contains an empty environment,
store and stack.
\begin{lstlisting}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{lstlisting}

Then the @drive@ function is used to evaluate
to a final state by iteratively apply @step@ on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete
abstract machines \cite{van2012systematic, van2010abstracting}.
An abstracting abstract machine implements computable abstract semantics that
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibly infinite,
to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Since each component of state is finite,
the abstracted machine-state space is finite, and computable.

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show how to instantiate
useful $k$-call-sensitive control-flow analysis.

@@

Similar to CESK machines, the machine state of AAM has a control expression,
an environment, a store and continuation, and in addition with a time-stamp.
But there are several notable differences between AAM's store and CESK machine's store.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus dereferencing addresses becomes non-deterministic.
Also, the store performs joining rather than overwriting when updating elements
of the store.
Third, the continuations are likewise allocated on store instead of formed a linked list,
and the continuation in a state then turns into a continuation address.

For clarity, we divide the store into a binding store @BStore@ and a continuation store @KStore@.
The binding store maps binding addresses to sets of closure values, and the continuation store maps
continuation addresses to sets of continuations.
We define a generic class @Store[K,V]@ that performs joining when updating elements
in a store. By parameterizing @Store[K,V]@ with @[BAddr, Storable]@ and
@[KAddr, Cont]@, we obtain @BStore@ and @KStore@, perspectively.

TODO(rephrase): Note that the store is monotonic, the store is always growing and never becoming
smaller. This property guarantees we eventually will find a fixed point.

@@

The co-domain of binding stores @Storable@ is the same
as we defined for CESK machines.
The co-domain of continuation stores @Cont@ is comprised of
a frame @Frame@ and a continuation address @KAddr@.
To mimic the runtime call stack, @KAddr@ plays the role of
representing the remaining stack frames.
TODO: talk about nondet of cont deref?

@@

As a consequence, the components of states are also changed:
the store is divided into binding store and continuation store;
the continuation stack becomes an address.
By dereferencing this address in a continuation store, we can retrieve the
actual transfer of controls. The definition of environment @Env@
remains the same.

@@

\subsubsection{Single-step Transition}

Since dereferencing an address becomes non-deterministic, our @atomicEval@
function also turns to be nondeterministic. Given an atomic expression @e@,
@atomicEval@ returns a set of storable values, i.e., closures, to the caller.
If the expression is just a @lambda@ term, then the returned set is a singleton.

@@

The structure of function @step@ is similar to the concrete CESK machine,
except non-determinism which makes @step@ return a list of reachable successor states.
In the beginning of @step@, we call function @tick@ to update the time.
Then we have two cases to consider:

@@

In this section, we describe a variant of AAM that leaves the stack to be
unbounded. That uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtime's
(so called pushdown control-flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section \ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return match
used for the next step of transformation.

@@

The state transition function @step@ shows below is still
non-deterministic, but the only non-determinism happens when dereferencing the callee @f@
from function applications @App(f, ae)@.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentioned for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Here we are facing a problem:
the underlying semantics of AAM is fundamentally non-deterministic.
But the evaluation context (the @Frame@ in our program) is not
non-deterministic, and the work list @todo@ actually implicitly handles the non-determinism,
if we simply refunctionalize the frames to functions, it would not helps toward
abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all the non-deterministic choices.
This step turns the @step@ function to be deterministic.
Likewise the @Frame@ saves the information of the caller in concrete executions,
we introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state which is called
@NDState@ now.
For the rest of the paper, to differentiate the two continuations, we would call
the first one as normal continuation, and the second one as non-deterministic
continuations.

@@

By atomically evaluating @f@, we obtain a set of closures.
The transition is deterministic to the first element of the closure set.
So we just prepare new environment, new store, and new frame list only for the
first closure in that set.
A new non-deterministic continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store,
the time and the new frame list.
We use the store before updating because for different closures they may form
different binding addresses @baddr@.
We also use the new frames, because all the closures at this fork point share
the same stack and return point.

In the second case shown in below we will see how @NDCont@ deal with
the non-determinism.

@@

But, since we have added non-determinism into state, an empty list of frames means
we have the reached the halt of \textit{one computation path}, and we should
determine the next state indicates by @NDCont@.
So we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we had tried all the closures of this fork point and should move the the
next fork point. Since the @NDCont@ is represented by a list and we always
append new element from its front, we are resuming to the most recent fork point by
popping up the front-most element.
\item Otherwise, we pop a closure from the set, then build new
environment and new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The non-determinism continuation @new_ndk@ is also updated by removal of that closure.
\end{itemize}

@@

The function @drive@ is also changed: there is no work list anymore, because all the
potentially unexplored states are embedded in to the continuation for
non-determinism.
The termination of the analysis is when we reach a @NDState@ that the expression
is atomic, and both the normal continuation and non-determinism continuation are
empty list. This corresponds to the case that @todo@ list is empty.

@@

Since there are two layers of continuations, as a result,
we obtain three mutually recursive functions, @drive_step@,
@continue@ and @ndcontinue@. The function @drive_step@ plays the same
role as before; @continue@ handles the evaluation context (@Frame@) of the analyzed
language, which the function is called from @drive_step@ when encountering an atomic
expression; @ndcontinue@ dispatches non-deterministic continuations @NDCont@,
and is invoked from @continue@ when the normal stack is empty.

@@

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation
  function.
\item Since our final target is abstract definitional interpreters while still
non-deterministic, start from this section, our abstract semantic artifact
returns a set of final values instead of collected states.
\item As we are getting close to the
abstract definitional interpreters, the name of @drive_step@ is updated to
@aeval@ (for \emph{abstract eval}).
\end{itemize}

For clarity, in this section, we first show how to refunctionalize the normal continuation and
non-deterministic continuation without collecting semantics.
Then adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is just changing the program to cache-passing style.
Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

To properly represented final values, we introduce the case class @VS@ which
contains a set of storable values, a time stamp, and a store.
\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
\end{lstlisting}
An instance of @VS@ represents the computational result of one path in
the non-deterministic evaluation.
The reason that we include a store is there might be a case two pathes have the
same values but different accumulated side effects such as memory allocations.
Since the whole computation is non-deterministic, the type of final result
values is a collection @Set[VS]@. For convenience, we define it as answer type @Ans@:
\begin{lstlisting}
type Ans = Set[VS]
\end{lstlisting}

@@

\textbf{Non-Determinism Abstraction.}
To handle the inevitable non-determinism, we defined a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

@@

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the non-deterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evaluation. \todo{refine}

@@

\subsection{Pushdown Control-Flow Analysis, Revisited} \label{pdcfarevisit}

>>>>

%% Title information
%\title{Pushdown Control Flow Analysis via Refunctionalization (Functional Pearl)}         %% [Short Title] is optional;
\title{Refunctionalization of Abstracting Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap Between Abstracting Abstract Machines and Abstracting Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and monadification (or un-CPS to direct-style). Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

\note{rest of intro goes here}

The rest of this paper is structured as follows: we begin by reviewing some background
necessary for this work in Section~\ref{background}, as well as introducing some of the basic
code structures used throughout the paper. We then address the main contribution of this
paper, which is the filling of the gap between small-step AAM and big-step ADI by developing a
series of systematic transformations. Those transformations are summarized here, with their
associated section:

\begin{itemize}
\item We show how \textbf{linearization} (Section~\ref{linear}) can be used to...
\item In Section~\ref{fusing}, we then present the \textbf{fusing} transformation, and present...
\item Section~\ref{disen} discusses \textbf{disentangling}, which shows...
\item \textbf{Refunctionalization} is shown in Section~\ref{refunc}, and brings a new understanding of pushdown control flow analysis. In this section, we present an abstract interpreter obtained from refunctionalizing a small-step abstracting abstract machine which naturally exhibits the pushdown control-flow property (as refunctionalization requires abstract execution to be sequentialized).
\item The last transformation we show is that of transforming a definitional abstract interpreter to a \textbf{direct-style} interpreter (Section~\ref{directstyle}).

These transformations are used throughout the paper, with refunctionalization and
defunctionalization of abstract interpreters playing important roles for the call stack of
the analyzed language (i.e., Scala)\note{right?}. All of this insight allows the revelation of
precisely \emph{why} small-step AAM does not have the pushdown control-flow property
(without extra effort), as well as explains why ADI naturally extends pushdown control-flow
property from its defining language.

We complete this paper by discussing related work in Section~\ref{sec:related}, followed by concluding thoughts in Section~\ref{sec:conclusion}.

% \subsection{Contributions}

@@

% We summarize our contributions as follows:

% \begin{itemize}
% \item The main contribution of this paper is the filling of the gap
%   between small-step AAM and big-step ADI by developing a series of
%   systematic transformations.
%   We show that the correspondence between abstract
%   machines and interpreters exists not only in concrete semantics artifacts,
%   but also in abstract semantics artifacts.
%   Galois connection?
%   \note{JD: Section number?}

% \item A new understanding of pushdown control flow analysis by
%   refunctionalization.
%   We present an abstract interpreter obtained from refunctionalizing
%   a small-step abstracting abstract machine.
%   This abstract interpreter naturally exhibits the pushdown control-flow property,
%   as the refunctionalization forces us to sequentialize the
%   abstract execution.
%   We also analyze its correctness, soundness, and termination.

% \item Further, we show that defunctionalization and refunctionalization
%   of abstract interpreters play important roles for the call stack of the analyzed language.
%   Given this insight, we reveal precisely \emph{why} small-step AAM has no pushdown
%   control-flow property (without extra efforts), as well as explain
%   why ADI naturally extends pushdown control-flow property from its
%   defining language.

% \end{itemize}

% \subsection{Outline}

% \note{TR: can this be merged with contributions? (they currently lack section refs)}
% The remainder of this paper is organized as follows.
% We begin by describing the target language we will analyze in Section~\ref{background},
% including a concrete CESK machine as small-step operational semantics for it.
% We continue with a review of abstracting abstract machines (AAM) as an abstract interpretation
% counterpart of CESK machines. Following this, Section~\ref{linear} shows a conversion from the
% nondeterministic choices of closures to another layer of continuations,
% therefore linearizing the abstract execution.
% Section~\ref{fusing} then discusses fusing the driver and step functions to a single
% function.
% We further identify in Section~\ref{disen} the first-order data type representations of
% continuations, and disentangle the code blocks for dispatching data types to separate
% functions.
% Section~\ref{refunc} then represents the refunctionalized abstract abstract machines.
% In this section, the continuations represented by data types become higher-order
% functions.
% In Section~\ref{directstyle}, we transform the refunctionalized AAM to two
% different direct styles: one implemented with monads, the other using
% delimited controls.

% Finally, we discuss related work in Section~\ref{sec:related}, followed by concluding thoughts in Section~\ref{sec:conclusion}.

\todo{Add section numbers to the figure}\note{TR: put this into a proper figure, add known danvy sequence
for concrete evaluators at the bottom (in grey or dashed lines)}
\note{JD: should these be abstracting abstract machines?}

@@

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because it exposes control transfer explicitly
and simplifies analysis~\cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use $\lambda$-calculus in administrative normal form (ANF),
which is a direct-style intermediate representation, as our target for clarity,
but without losing simplicity and generality.
Although we only show the core calculus language, it can be easily extended
to support recursive bindings (such as @letrec@), conditionals, primitive types, and
operations on primitive types. These cases would be trivial to implement,
so we elide them in this paper.
The transformations we will show in the rest of this paper
also work on abstract machines for plain direct-style $\lambda$-calculus languages.

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language
in ANF~\cite{flanagan1993essence}.

@@

In ANF, an expression is either an atomic expression or a @let@ expression.
A restriction exists which states that all function applications must be administrated
within a @let@ expression and then bound to a variable name under the current environment.
Both the operator and operand of function applications are atomic expressions.
An atomic expression $ae$ is either a variable or a literal @lambda@ term, either of which
can be evaluated in a single step.
We also assume that all the variable names in the program are unique.

@@

The CESK machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
The CESK machine models program execution as state transitions in a small-step fashion.
As its name suggests, a machine state has four components:
1) \textit{Control} is the expression currently being evaluated.
2) \textit{Environment} is a map that contains the address of a variable in the lexical scope.
3) \textit{Store} models the heap of a program as a map from addresses to values.
The address space consists of numbers (0-indexed).
In our toy language, the only category of value is a closure, i.e., a function paired with an environment.
4) \textit{Continuation} represents the program stack. \todo{explain more on stack}

The Scala representations for the components of CESK machine are as follows:

@@

It is worth noting that the continuation class @Kont@ is defined as a list of
frames, where a frame can be considered an evaluation context in reduction semantics.
We represent frames using the @Frame@ class, which stores the information of a single
call-site, i.e., the information that can be used to resume the interrupted computation.
A @Frame@ constitutes a variable name @x@ to be bound later, a control expression
to which the program may resume, and its environment. \note{JD: someone check this and make sure I got it right}
\todo{mention that frame list vs KArg/KFun}

\subsubsection{Single-Step Transition}
Before describing how the machine evaluates expressions, we must first define several helper
functions. As mentioned in Section~\ref{anfsyntax}, atomic expressions are either a variable
or a literal @lambda@ term. As such, the atomic expression evaluator @atomicEval@ handles
these two cases and evaluates atomic expressions to closures in a straightforward way.
The @alloc@ function generates a fresh address, and always allocates a unique integer
in the domain of @store@.
The @isAtomic@ function is used as a predicate to determine if the expression is atomic.

@@

We can now faithfully describe the state transition function @step@,
which when given a machine state, determines its successor state.

@@

As shown and previously discussed, we examine the only two cases which the state may be.

\textbf{\texttt{Let} expression.} In the first @case@ statement shown in the previous code,
the control of the current state matches as a @Let@ expression, with its right-hand side a
function application.
By calling the @atomicEval@ evaluator, we obtain the closure for which the callee @f@ stands.
The successor state's control then transfers to the @body@ expression of the closure
with an updated environment and store. The new environment is extended
from the closure's environment and mapped from @v@ to a fresh address @addr@.
The new store is extended with @addr@ mapping to the value of @ae@,
which in turn is evaluated from @atomicEval@.
Finally, a new frame @frame@ is pushed onto the stack @k@, where
@frame@ contains the variable name @x@ at the left-hand position of the @Let@,
the body expression of @Let@, and the lexical environment of the body expression.

\textbf{Atomic expression.} If the state is not a @Let@ expression, then it must be
an atomic expression, as seen in the above code. In this scenario,
we begin by extracting the top frame of all available continuations.
The control (i.e., an atomic expression) of the current state is the evaluated term
that is being bound to the variable @x@ from the top frame.
The environment and store are updated with @x@ mapping to the closure value of @ae@.
Finally, the successor state is transferred to expression @e@ from the top frame,
which is the body of a @Let@ expression, with the updated environment, store, and
the rest of the stack @ks@.

@@

To run the program, we first use the @inject@ function (below) to construct an initial machine
state given an expression @e@. The initial state contains an empty environment,
store, and stack.
\begin{lstlisting}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{lstlisting}

The @drive@ function is then used to evaluate
to a final state by iteratively applying @step@ on the current state until a state is reached
ion which the control is an atomic expression and the continuation stack is empty.
Naturally, we can then extract the value from the final state at last.

@@

\subsection{Abstracting Abstract Machines} \label{aam}
Abstracting abstract machines (AAM) is a systematic methodology that derives sound
abstract interpreters for higher-order functional languages from concrete
abstract machines \cite{van2012systematic, van2010abstracting}.
An abstracting abstract machine implements computable abstract semantics which
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibly infinite,
the key insight of AAM approach when analyzing programs is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Since each component of state is finite,
the abstracted machine-state space is also finite, and therefore computable.

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show how to instantiate
useful $k$-call-sensitive control flow analysis.

@@

Similar to CESK machines, the machine state of AAM has a control expression,
an environment, a store, and continuation, as well as a timestamp.
However, there are several notable differences between AAM's store and CESK machine's store.
In AAM, the store maps addresses to sets of values; it stores all possible values
for a particular address. As such, dereferencing addresses becomes nondeterministic.
Also, the store performs \emph{joining}, rather than overwriting, when updating elements.
Furthermore, the continuations are likewise allocated on the store instead of formed into a
linked list, and the continuations are in a state which then turns into a continuation
address.

For clarity, we divide the store into two separate stores: the binding stores @BStore@, and
the continuation store @KStore@.
The binding store maps binding addresses to sets of closure values, whereas the continuation
store maps continuation addresses to sets of continuations.
We then define a generic class @Store[K,V]@ that performs joining when updating elements
in a store (below). By parameterizing @Store[K,V]@ with @[BAddr, Storable]@ and
@[KAddr, Cont]@, we obtain @BStore@ and @KStore@, respectively.

We note that the \note{store? update function? can a store be monotonic?} is monotonic; it continuously grows and never shrinks. This property
guarantees that in analyzing the \note{update function? store? pick the right word}, a fixed point can always be reached.

@@

The codomain of all binding stores @Storable@ is the same
as previously defined for CESK machines.
The codomain of continuation stores @Cont@, on the other hand, is comprised of
a @Frame@ object and a continuation address @KAddr@.
To mimic the runtime call stack, @KAddr@ plays the role of
representing the remaining stack frames.
\todo{talk about nondet of cont deref?}

@@

As a consequence, the components of states are also changed:
the store is divided into binding stores and continuation stores;
the continuation stack becomes an address.
By dereferencing this address in a continuation store, we can retrieve the
actual transfer of controls. The definition of environment @Env@
remains the same.

@@

\subsubsection{Single-Step Transition}

Since dereferencing an address becomes nondeterministic, our @atomicEval@
function also turns to be nondeterministic. Given an atomic expression @e@,
@atomicEval@ returns a set of storable values, i.e., closures, to the caller.
If the expression is just a @lambda@ term, then the returned set is a singleton.

@@

The structure of function @step@ is similar to the concrete CESK machine,
except nondeterminism which makes @step@ return a list of reachable successor states.
In the beginning of @step@, we call function @tick@ to update the time.
Then we have two cases to consider:

@@

In this section, we describe a variant of AAM that leaves the stack to be
unbounded. That uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtime's
(so called pushdown control flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section \ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return match
used for the next step of transformation.

@@

The state transition function @step@ shows below is still
nondeterministic, but the only nondeterminism happens when dereferencing the callee @f@
from function applications @App(f, ae)@.

@@

In Danvy's paper \textit{Defunctionalized Interpreters for Programming
  Languages}, he mentioned for deterministic languages, a reduction semantics
is a structural operational semantics in continuation style, where the reduction
context is a defunctionalized continuation \cite{Danvy:2008:DIP:1411204.1411206}.
Here we are facing a problem:
the underlying semantics of AAM is fundamentally nondeterministic.
But the evaluation context (the @Frame@ in our program) is not
nondeterministic, and the work list @todo@ actually implicitly handles the nondeterminism,
if we simply refunctionalize the frames to functions, it would not helps toward
abstract definitional interpreters.

In order to address this problem, our first step of transformations
is to linearize all the nondeterministic choices.
This step turns the @step@ function to be deterministic.
Likewise the @Frame@ saves the information of the caller in concrete executions,
we introduce another layer of controls, and define a case class @NDCont@
that saves the information at a fork point when we have multiple target closures.
We also add a new field @ndk@ to the definition of state which is called
@NDState@ now.
For the rest of the paper, to differentiate the two continuations, we would call
the first one as normal continuation, and the second one as nondeterministic
continuations.

@@

By atomically evaluating @f@, we obtain a set of closures.
The transition is deterministic to the first element of the closure set.
So we just prepare new environment, new store, and new frame list only for the
first closure in that set.
A new nondeterministic continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store,
the time and the new frame list.
We use the store before updating because for different closures they may form
different binding addresses @baddr@.
We also use the new frames, because all the closures at this fork point share
the same stack and return point.

In the second case shown in below we will see how @NDCont@ deal with
the nondeterminism.

@@

But, since we have added nondeterminism into state, an empty list of frames means
we have the reached the halt of \textit{one computation path}, and we should
determine the next state indicates by @NDCont@.
So we have a pattern matching on @ndk@:
\begin{itemize}
  \item If the closure set is an empty set,
then we had tried all the closures of this fork point and should move the the
next fork point. Since the @NDCont@ is represented by a list and we always
append new element from its front, we are resuming to the most recent fork point by
popping up the front-most element.
\item Otherwise, we pop a closure from the set, then build new
environment and new store for the body expression of the closure;
we use the same frame list and time that are copied from the fork point rather
than current one.
The nondeterminism continuation @new_ndk@ is also updated by removal of that closure.
\end{itemize}

@@

The function @drive@ is also changed: there is no work list anymore, because all the
potentially unexplored states are embedded in to the continuation for
nondeterminism.
The termination of the analysis is when we reach a @NDState@ that the expression
is atomic, and both the normal continuation and nondeterminism continuation are
empty list. This corresponds to the case that @todo@ list is empty.

@@

Since there are two layers of continuations, as a result,
we obtain three mutually recursive functions, @drive_step@,
@continue@ and @ndcontinue@. The function @drive_step@ plays the same
role as before; @continue@ handles the evaluation context (@Frame@) of the analyzed
language, which the function is called from @drive_step@ when encountering an atomic
expression; @ndcontinue@ dispatches nondeterministic continuations @NDCont@,
and is invoked from @continue@ when the normal stack is empty.

@@

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation
  function.
\item Since our final target is abstract definitional interpreters while still
nondeterministic, start from this section, our abstract semantic artifact
returns a set of final values instead of collected states.
\item As we are getting close to the
abstract definitional interpreters, the name of @drive_step@ is updated to
@aeval@ (for \emph{abstract eval}).
\end{itemize}

For clarity, in this section, we first show how to refunctionalize the normal continuation and
nondeterministic continuation without collecting semantics.
Then adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is just changing the program to cache-passing style.
Finally, we examine how computable pushdown control flow analysis is established
through refunctionalization and caching.

@@

To properly represented final values, we introduce the case class @VS@ which
contains a set of storable values, a time stamp, and a store.
\begin{lstlisting}
case class VS(vals: Set[Storable], time: Time, store: BStore)
\end{lstlisting}
An instance of @VS@ represents the computational result of one path in
the nondeterministic evaluation.
The reason that we include a store is there might be a case two pathes have the
same values but different accumulated side effects such as memory allocations.
Since the whole computation is nondeterministic, the type of final result
values is a collection @Set[VS]@. For convenience, we define it as answer type @Ans@:
\begin{lstlisting}
type Ans = Set[VS]
\end{lstlisting}

@@

\textbf{NonDeterminism Abstraction.}
To handle the inevitable nondeterminism, we defined a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

@@

\textbf{First Version of Evaluation Function.}
Since our first version of @aeval@ just implements the nondeterministic
evaluation, and given the defined @nd@, the evaluation function @aeval@ can be easily
defined. For the case that @e@ is a @let@ expression,
the overall idea is to perform a two step depth-first evaluation. \todo{refine}

@@

\subsection{Pushdown Control Flow Analysis, Revisited} \label{pdcfarevisit}
